---
title: "Readme - Getting and Cleaning Data Course Project"
author: "J2"
date: "February 24, 2016"
output: html_document
---
#Getting and Cleaning Data Course Project

##Overview
Human Activity Recognition database comprises the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors.  The experiments were carried out with a group of 30 volunteers, each person performed six activities (WALKING, WALKING UPSTAIRS, WALKING DOWNSTAIRS, SITTING, STANDING, LAYING), while wearing a smartphone (Samsung Galaxy S II) on the waist.  Using its embedded accelerometer and gyroscope, the researchers captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz.  The obtained dataset was randomly partitioned into two sets, where 70% (7352 observations) of the volunteers were selected for generating the training data and 30% (2947 observations) selected for the test data.

##Project Objectives
The purpose of this project is to translate the raw data which was downloaded from the Human Activity Recognition Using Smartphones Data Set Project and prepare two tidy data sets to be used for future analysis, in accordance with the course project instructions listed below: 
  1.  Merge the training and the test sets to create one data set.
  2.  Extract only the measurements on the mean and standard deviation for each measurement.
  3.  Use descriptive activity names to name the activities in the data set
  4.  Appropriately label the data set with descriptive variable names.
  5.  From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

##Prerequisites
The project requires the following R package libraries as follows.
```{r}
library(dplyr)
library(data.table)
library(downloader)
```

## Download Data
The following code creates a new directory, run_analysis (if not already created), then downloads and unzips the data. 
```{r}
setwd("C:\\Users\\John\\Documents\\Dropbox\\Online Learning\\Data Science\\Data Science Specialization\\Getting & Cleaning Data\\Code")
if (!file.exists("run_analysis")){
  dir.create("run_analysis")
  setwd("./run_analysis")
  url = "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
  download(url, dest="dataset.zip", mode="wb") 
  unzip ("dataset.zip")
  setwd("UCI HAR Dataset")
  
} else {
  setwd("run_analysis/UCI HAR Dataset")
}
```

## Read Data
This code reads the key files required for analysis into tables for further analysis.  The files are as follows:
* features.txt - lists all 561 features measured during the experiment.
* activity_lables.txt - lists a number and a label for each of the 6 activities observed.
* subject_test.txt - lists the subject identifiers for each of the subjects observed in the test dataset.
* X_test.txt - a matrix specifying the values for each of the 561 variables measured over the 2947 observations in the test dataset
* y_test.txt - a vector listing the activity associated with each of the 2947 observations in the test dataset
* subject_train.txt - lists the subject identifiers for each of the subjects observed in the training  dataset.
* X_train.txt - a matrix specifying the values for each of the 561 variables measured over the 7352 observations in the training dataset
* y_train.txt - a vector listing the activity associated with each of the 7352 observations in the training dataset

```{r}
features        <- read.table("features.txt")
activity_labels <- read.table("activity_labels.txt",stringsAsFactor=FALSE)
test_subjects   <- read.table("test/subject_test.txt")
test            <- read.table("test/X_test.txt")
test_activity   <- read.table("test/y_test.txt")
train_subjects  <- read.table("train/subject_train.txt")
train           <- read.table("train/X_train.txt")
train_activity  <- read.table("train/y_train.txt")
```

## Combine Subject Data
The following code combines the subject data from both the test and training datasets, into a single dataframe.  
```{r}
subjects <-rbind(test_subjects, train_subjects)
names(subjects) <- "subject"
```

## Combine Activity Data
The following code combines the activity data  from both the test and training datasets, into a single dataframe.
```{r}
activities <- rbind(test_activity, train_activity)
names(activities) <- "activities"
```

## Combine Measurement Data
The following code combines the measurement data from both the test and training datasets, into a single dataframe. 
```{r}
measurements <- rbind(test, train)
names(measurements) <- tolower(features$V2)
```

## Select Mean and Standard Deviation Variables
The following code Selects the 66 mean and standard deviation variables from measurement data.  Note, 20 variables measuring meanFreq and angle gravityMean have been omitted from the selection since these variables are derived from the 66 variables selected.
```{r}
measurements <- measurements[,grepl("(mean\\(\\)|std\\(\\))", names(measurements))]
```

## Assign Descriptive Variable Names
The following code assigns descriptive variable names to the selected features.
```{r}
names(measurements) <- names(measurements) %>%
  sub("^t", "time-", .)   %>%
  sub("^f", "frequency-", .)   %>%
  gsub("\\.","",.)  %>%
  gsub("body", "body-", .) %>%
  gsub("gravity", "gravity-", .) %>%  
  gsub("acc", "accelerometer-", .) %>%  
  gsub("gyro", "gyroscope-", .) %>%
  gsub("mag", "magnitude-", .) %>%
  gsub("std\\(\\)", "standard-deviation", .) %>%
  gsub("mean\\(\\)", "mean", .) 
```  

## Combine Subject, Activity and Measurement Data
This code combines the subject, activity, and measurement data into a single dataframe.  In addition, the activity labels are merged into the dataframe, and the activity numbers are deleted.
```{r}
analysis <- cbind(subjects, activities, measurements)
analysis <- merge(analysis, activity_labels, by.x = "activities", by.y = "V1")
analysis$activities <- NULL
```

# Summarize Data
The following code summarizes the data by subject and activity(column v2) and uses the summarise_each function to calculate the mean of the 66 variables selected above.  Lastly, the activity label column is appropriately renamed in both the analysis and analysis summary dataframes.
```{r}
analysis_summary <- analysis %>% group_by(subject, V2) %>% summarise_each(funs(mean))
names(analysis)[names(analysis) == "V2"] <- "activity-labels"
names(analysis_summary)[names(analysis_summary) == "V2"] <- "activity-labels"
```
# Write Files
The following code writes both files in .csv format.
```{r}
write.table(analysis, file = "run_analysis.txt", sep="\t", row.names = FALSE)
write.table(analysis_summary, file = "run_analysis_summary.txt", sep="\t", row.names = FALSE)
'''
